---
title: "Predict CD8 immune phenotypes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predict CD8 immune phenotypes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(cd8ippred)
library(Biobase)
library(dplyr)
library(tidyr)
library(ggplot2)
```

This is an example of how to apply CD8 immune phenotype model to data.

**Warning**: Only apply this to RNA-Seq data processed with Roche-internal
RNA-Seq pipeline (e.g., processed EDIS data).
When in doubt, compare your data distribution to the example data.

We are using simulated data sampled from the training set.
First we need Variance-stabilizing transformation.

```{r "vst transformation"}
cd8ip_simulated_counts <- load_simulated_counts()
vst_m <- compute_vst(exprs(cd8ip_simulated_counts))
```

Create a `data.frame`.
```{r}
# set rownames to symbol names where available;
# the model uses only a subset of genes with symbols available.
rownames(vst_m) <- ifelse(
    is.na(fData(cd8ip_simulated_counts)$symbol) | fData(cd8ip_simulated_counts)$symbol == "",
    rownames(vst_m),
    fData(cd8ip_simulated_counts)$symbol
)

# make rownames unqiue; the model uses only a subset of genes
# without symbol duplication issues
rownames(vst_m) <- make.names(rownames(vst_m), unique = TRUE)

# transform to a data.frame()
vst_df <- as.data.frame(t(vst_m))
# data preview
vst_df[1:6, 1:6]
```

Predict immune phenotypes
```{r}
pred <- predict_cd8ip(vst_df)
pred$truth <- cd8ip_simulated_counts$CD8IMMPH |>
  tolower() |>
  # we convert truth to a factor for yardstick ROC computations
  factor()
```

Prediction quality
```{r}
# values above 6 should be interpreted with caution
summary(pred$n_out_of_range)

# low-confidence predictions are set to NA
summary(is.na(pred$predicted.class))
```

## Confusion matrix
```{r "confusion matrix"}
with(
  pred,
  table(truth, predicted.class)
)
```

View data.frame with predictions
```{r "predictions"}
head(pred)
```

## AUC scores
Please note, this is **not real data**. These AUC scores are only an example.

```{r "AUC data.frame"}
auc_df <- pred |>
  filter(!is.na(truth)) |>
  pivot_longer(starts_with("prob."), names_prefix = "prob.", names_to = ".level", values_to = "prob") |>
  mutate(positive = as.factor(truth == .level)) |>
  group_by(.level) |>
  yardstick::roc_auc(positive, prob, event_level = "second") |>
  rename(AUC = .estimate, class = .level) |>
  select(class, AUC) %>%
  mutate(AUC_text = paste0("AUC=", round(AUC, 3)))

select(auc_df, -AUC_text)
```


ROC curve
```{r "plot ROC", fig.width=5, fig.height=4}
pheno_col <- c(
  inflamed = "#d95f02",
  excluded = "#1b9e77",
  desert = "#7570b3"
)

pred |>
  filter(!is.na(truth)) |>
  # column names should match levels of the truth factor
  yardstick::roc_curve(truth, all_of(paste0("prob.", levels(pred$truth)))) |>
  mutate(class = .level) |>
  ggplot() +
  aes(x = 1 - specificity, y = sensitivity, col = class) +
  geom_path() +
  geom_abline(lty = 3) +
  coord_equal() +
  geom_text(aes(label = AUC_text), x = 0.55, y = 0.1, data = auc_df, show.legend = FALSE) +
  facet_wrap(~class, ncol = 2) +
  scale_color_manual(values = pheno_col) +
  theme_bw()
```

<details>
<summary>Session info</summary>
```{r "session info"}
devtools::session_info()
```
</details>
